---
title: "ID271 Midterm"
author: "JCO"
date: "Due Feb 22, 2026 at 11:59 PM (EST)"
output: 
  html_document:
    toc_float: true
    toc: true
---

### Methods

# Packages

```{r, message=F, include=FALSE}
library(foreign)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(stats)
library(mgcv)
library(table1)
library(splines)
library(nlme)
library(gt)
library(lubridate)
library(gplots)
library(MASS)
library(tidyverse)
library(nlme)
```

```{r, include=FALSE}
nas <- readRDS("nas_bp.rds")
nas$DATE <-as.Date(nas$DATE, format="%m/%d/%Y")
nas$WKDAY <- factor(nas$WKDAY, levels = 1:7) 
```

```{r}
summary(nas$SBP)
hist(nas$BMI)
hist(nas$AGE)
hist(nas$SBP)
hist(nas$BC24H)
hist(nas$TEMPC24H)
hist(nas$RHUM24H)
hist(nas$NEDUC)


hist(log(nas$SBP))

```

```{r sameday_model_sp_v1}
sameday_sp <- gamm(log(SBP) ~ s(BMI)+ s(BC24H) + WARMCOLD + s(AGE) + s(TEMPC24H)  + s(RHUM24H) + as.factor(RACE) + TWODRINK + CHD + DIABETE + as.factor(SMK) + as.factor(FBG) + s(NEDUC) + STATIN + SINETIME + COSTIME + as.factor(WKDAY) + as.factor(VISIT), na.action = na.omit, random = list(ID =  ~ 1), family = gaussian(), data = nas)
plot(sameday_sp$gam,  scale = 0)
cor(nas$WARMCOLD, nas$SINETIME)
cor(nas$WARMCOLD, nas$COSTIME)
summary(sameday_sp$lme)
```

#coldwarm should be removed, highly collinear with COSTIME (representative of WINTER and SUMMER); however need to keep in the model because of one of the questiosn
```{r sameday_model_sp_v2}
sameday_sp2 <- gamm(log(SBP) ~ BC24H + BMI + WARMCOLD + s(AGE) + TEMPC24H  + RHUM24H + as.factor(RACE) + TWODRINK + CHD + DIABETE + as.factor(SMK) + as.factor(FBG) + NEDUC + STATIN + SINETIME + COSTIME + as.factor(WKDAY)+  as.factor(VISIT), na.action = na.omit, random = list(ID =  ~ 1 + VISIT), family = gaussian(), data = nas)
plot(sameday_sp2$gam,  scale = 0)
summary(sameday_sp2$lme)
```


```{r sameday_model}
hist(sameday_sp2$lme$coef$random$ID[,2]+sameday_sp2$lme$coef$fixed[2], main="Distribution of BC24H Slopes by ID", xlab="BC24H on log SBP")

ID<-unique(nas$ID)
 
plot(ID,sameday_sp2$lme$coef$random$ID[,2]+sameday_sp2$lme$coef$fixed[2] )
abline(lsfit(ID,(sameday_sp2$lme$coef$random$ID[,2]+sameday_sp2$lme$coef$fixed[2])),col=3)
```
 Are there any subgroups that are particularly vulnerable to the effect of same-day BC exposure on SBP?
 - No (out of age, race, BMI, statin, smoking ) YES CHD
 Does diabetes status modify the effect of same-day BC exposure on SBP? If so, provide the effect estimates (with 95% CI) between BC and SBP among diabetics and non-diabetics, respectively.
 - No 

```{r}
sameday_sp3 <- gamm(log(SBP) ~ BC24H*AGE + BMI + WARMCOLD + TEMPC24H  + RHUM24H + as.factor(RACE) + TWODRINK + CHD + DIABETE + as.factor(SMK) + as.factor(FBG) + NEDUC + STATIN + SINETIME + COSTIME + as.factor(WKDAY)+  as.factor(VISIT), na.action = na.omit, random = list(ID =  ~ 1 + VISIT), family = gaussian(), data = nas)
summary(sameday_sp3$lme) #not significant 

sameday_sp4 <- gamm(log(SBP) ~ BC24H*RACE + BMI + WARMCOLD + s(AGE) + TEMPC24H  + RHUM24H + TWODRINK + CHD + DIABETE + as.factor(SMK) + as.factor(FBG) + NEDUC + STATIN + SINETIME + COSTIME + as.factor(WKDAY)+  as.factor(VISIT), na.action = na.omit, random = list(ID =  ~ 1 + VISIT), family = gaussian(), data = nas)
summary(sameday_sp4$lme) #not significant 

sameday_sp5 <- gamm(log(SBP) ~ BC24H*DIABETE + BMI + WARMCOLD + s(AGE) + TEMPC24H  + RHUM24H + as.factor(RACE) + TWODRINK + CHD + DIABETE + as.factor(SMK) + as.factor(FBG) + NEDUC + STATIN + SINETIME + COSTIME + as.factor(WKDAY)+  as.factor(VISIT), na.action = na.omit, random = list(ID =  ~ 1 + VISIT), family = gaussian(), data = nas)
plot(sameday_sp5$gam,  scale = 0)
summary(sameday_sp5$lme) #diabetes not significant 

sameday_sp6 <- gamm(log(SBP) ~ BC24H*BMI + WARMCOLD + BMI + s(AGE) + TEMPC24H  + RHUM24H + as.factor(RACE) + TWODRINK + CHD + DIABETE + as.factor(SMK) + as.factor(FBG) + NEDUC + STATIN + SINETIME + COSTIME + as.factor(WKDAY)+  as.factor(VISIT), na.action = na.omit, random = list(ID =  ~ 1 + VISIT), family = gaussian(), data = nas)
plot(sameday_sp6$gam,  scale = 0)
summary(sameday_sp6$lme) #ns

sameday_sp7 <- gamm(log(SBP) ~ BC24H*STATIN + WARMCOLD + BMI + s(AGE) + TEMPC24H  + RHUM24H + as.factor(RACE) + TWODRINK + CHD + DIABETE + as.factor(SMK) + as.factor(FBG) + NEDUC + BMI + SINETIME + COSTIME + as.factor(WKDAY)+  as.factor(VISIT), na.action = na.omit, random = list(ID =  ~ 1 + VISIT), family = gaussian(), data = nas)
plot(sameday_sp7$gam,  scale = 0)
summary(sameday_sp7$lme) #ns

sameday_sp8 <- gamm(log(SBP) ~ BC24H*CHD + WARMCOLD + s(AGE) + BMI + TEMPC24H  + RHUM24H + as.factor(RACE) + TWODRINK + CHD + DIABETE + as.factor(SMK) + as.factor(FBG) + NEDUC + STATIN + SINETIME + COSTIME + as.factor(WKDAY)+  as.factor(VISIT), na.action = na.omit, random = list(ID =  ~ 1 + VISIT), family = gaussian(), data = nas)
plot(sameday_sp8$gam,  scale = 0)
summary(sameday_sp8$lme) #SIGNIFICANT

sameday_sp8 <- gamm(log(SBP) ~ BC24H*as.factor(SMK) + WARMCOLD + s(AGE) + CHD + BMI + TEMPC24H  + RHUM24H + as.factor(RACE) + TWODRINK + CHD + DIABETE + as.factor(FBG) + NEDUC + STATIN + SINETIME + COSTIME + as.factor(WKDAY)+  as.factor(VISIT), na.action = na.omit, random = list(ID =  ~ 1 + VISIT), family = gaussian(), data = nas)
plot(sameday_sp8$gam,  scale = 0)
summary(sameday_sp8$lme) #ns
```













