---
title: "Association of Same-Day Ambient Fine Particulate Matter with Mortality"
author: "Anton Roche, Grace Gaultier, and Joel Chavez Ortiz"
date: "ID 271 - Midterm"
output: 
  html_document:
    toc_float: true
    toc: true
---

```{css, echo=FALSE}
.hanging-indent p {
  padding-left: 2em;
  text-indent: -2em;}
.side-by-side {
  display: flex;
  gap: 40px;
  align-items: flex-start;
}
.side-by-side .text {
  flex: 1;
}
.side-by-side .table {
  flex: 1.2;
}
```


```{r packages, message=F, include=FALSE}
library(foreign)
library(stats)
library(mgcv)
library(splines)
library(nlme)
library(table1)
library(lubridate)
library(gplots)
library(dplyr)
library(ggplot2)
library(MASS)
library(gt)
```

```{r data, include=FALSE}
hou_pm <- read.csv("hou_pm.csv")
hou_pm$date <-as.Date(hou_pm$date, format="%m/%d/%Y")
hou_pm$time<-1:dim(hou_pm)[1]
hou_pm$dow <- factor(hou_pm$dow, levels = 1:7) 
```

### Background

Emissions from both natural and anthropogenic sources can create particulate matter (PM) small enough to contribute to adverse health effects in humans (e.g., damage to lung function, asthma) and contribute to environmental damage (e.g., acid rain effects, disruption of nutrient balance in soil and rivers)$\text{.}^{1}$ PM with a diameter of 2.5 micrometers or smaller $\text{PM}_{2.5}$ are considered fine inhalable particles that can penetrate deeper into the lungs than larger PM. Houston is a large U.S. city with a large urban population. High PM emissions from vehicles, industry, and other sources may pose a risk to population health. An observational time-series analysis was conducted to further understand the association between same-day ambient concentrations of $\text{PM}_{2.5}$ and daily mortality.

### Methods

Using historical records data from Houston between January 1, 1999 and December 31, 2006, daily deaths, $\text{PM}_{2.5}$ concentration, and meteorologic were extracted to analyze the association between $\text{PM}_{2.5}$ and three outcomes of interest: all-cause mortality, cardiovascular (CVD) disease-related mortality, and respiratory disease-related mortality. Days in our study period without an average $\text{PM}_{2.5}$ were excluded.

#### Study Analysis

The association between same-day concentrations of $\text{PM}_{2.5}$ and the three outcomes of interest was estimated fitting generalized additive models (GAM) with a quasi-Poisson regression. When possible, relative risk (RR) was estimated between $\text{PM}_{2.5}$ and our outcomes if they seemed to have a linear relationship. For the CVD outcome, the relationship between $\text{PM}_{2.5}$ and CVD related mortality did not have a linear relationship when flexibly modeled, and thefore a spline was fit for $\text{PM}_{2.5}$ in the CVD outcome model. In this analysis, a quasi-Poisson distribution was used. A quasi-Poisson model provides an advantage over a standard Poisson regression in that it allows for data to be overdispersed. In our data, the mean did not equal the variance for any of the outcomes, signifying overdispersion and the need for the quasi-Poisson distribution. The adjusted GAMs included relative humidity (RH), day of the week (dow), average daily temperature (TEMP), and days since the start of the study (time) as relevant covariates. Time was modeled with a natural spline using 5 degrees of freedom for each of the 8 years of the study (40 degrees of freedom) to account for trends in seasonality and variation between years. 

Penalized splines were initially fit for relative humidity (RH) and temperature. Through the flexible modeling, RH and temperature were found to have linear relationships with the outcomes of interest, so the models were refit with RH and temperature as linear predictors. Splines were fit with cubic regression smoothing functions generated from restricted maximum likelihood (REML) for more stable parameter estimation and less overfitting. Analysis of variance (ANOVA) tests were conducted for each outcome, testing the null hypothesis that the model without temperature was sufficient, and the alternative hypothesis that temperature is necessary in the model.


A generic model for all three outcomes is below.   

$$\text{log}(\mathbb{E}[Y_{t}]) = \beta_{0} +f_{1}(PM_{t})+f_{2}(Time_{t})+\beta_{1}Temp_{t}+\beta_{2}RH_{t}+\beta_{3}Dow_{t}$$

Where $Y_t$ is the count of mortality (total, CVD, or respiratory) on day t since the beginning of the study, RH is the same day average relative humidity (%), Temp is the same day average temperature (Fahrenheit), $\text{PM}_{2.5}$ is the same day average fine particulate matter concentration (in $\mu g/m^3$), and day of week (dow) was included as an indicator variable in this model.

### Results

::::: side-by-side
::: text
Over the 8-year study period, 2,523 days had data on average same-day $\text{PM}_{2.5}$ concentration (86.34%; 2,253/2,922 days). The mean for daily all-cause, CVD, and respiratory mortality were 50.26 (variance: 62.56, SD: 7.91), 14.89 (variance: 16.70, SD: 4.09), and 3.99 (variance: 4.67, SD: 2.16), respectively (Table 1). The data for number of daily deaths by mortality outcome is right-skewed (Figure 1). Over the study period, there were a total of 126,802 recorded deaths from any cause. Of these deaths, 29.6% (37,578/126,802) were from CVD and 8% (10,066/126,802) were from respiratory disease. The median concentration of $\text{PM}_{2.5}$ was 12 $\mu g/m^3$, with the highest single day concentration being 53 $\mu g/m^3$. Average temperature over the 8 year period was 69.9° with average relative humidity being 73.08%.
:::

::: table
```{r table1, echo=FALSE}
hou_pm<-na.omit(hou_pm)
label(hou_pm$tot)  <- "All-cause mortality, daily"
label(hou_pm$cvd)  <- "Cardiovascular mortality, daily"
label(hou_pm$resp) <- "Respiratory mortality, daily"
label(hou_pm$pm25) <- "PM2.5, µg/m³"
label(hou_pm$TEMP) <- "Temperature, °F"
label(hou_pm$RH)   <- "Relative Humidity, %"

table1(~ tot + cvd + resp + pm25 + TEMP + RH,  data = hou_pm, caption = "<span style='color: black; font-weight: bold;'>Table 1. Descriptive Statistics of Daily Mortality, Daily Particulate Matter, and Weather, Houston 1999-2006", overall = "Descriptive Statistic Values</span>",
       render.continuous = function(x) {
         outc <- attr(x, "label")
         if (outc %in% c("All-cause mortality, daily", "Cardiovascular mortality, daily","Respiratory mortality, daily")) {
        c(" " = "", "Mean (SD)" = sprintf("%.2f (%.2f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE)),
           "Variance" = sprintf("%.2f", var(x, na.rm = TRUE)),
           
           "Min, Median, Max" = sprintf("%.0f, %.0f, %.0f", min(x, na.rm = TRUE),  median(x, na.rm = TRUE),  max(x, na.rm = TRUE)), "~Total Deaths, 1999-2006" = format(sum(x, na.rm = TRUE), big.mark=","))                            
         }
        else {
            c(" " = "", "Mean (SD)" = sprintf("%.2f (%.2f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE)),
            "Min, Median, Max" = sprintf("%.1f, %.0f, %.0f", min(x, na.rm = TRUE),  median(x, na.rm = TRUE),  max(x, na.rm = TRUE))) 
        }
       })

```
:::
:::::

```{r histograms, echo=FALSE}
par(mfrow = c(1, 3), oma = c(4, 4, 4, 0), mar = c(2, 2, 2, 1)) 
hist(hou_pm$tot, main ="All-Cause Mortality",)
hist(hou_pm$cvd,main ="CVD Mortality")
hist(hou_pm$resp, main ="Respiratory Mortality")
mtext("Figure 1. Histograms of Daily Mortality, Houston 1999-2006", side = 3, line = 1, outer = TRUE, cex = 1.2, font = 2)
mtext("Number of Deaths ", side = 2, line = 1, outer = TRUE, cex = 1.0, font = 2)
mtext("Frequency", side = 1, line = 1, outer = TRUE, cex = 1.0, font = 2)
```

```{r models, include=FALSE}
mod_tot<-gam(tot ~ pm25 + s(time,bs = "cr",fx=TRUE,k=41)+TEMP+ RH + as.factor(dow), family=quasipoisson, data=hou_pm, method="REML")

mod_cvd<-gam(cvd ~ s(pm25) + s(time,bs = "cr",fx=TRUE,k=41)+TEMP+ RH + as.factor(dow), family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML")

mod_resp<-gam(resp ~ pm25 + s(time,bs = "cr",fx=TRUE,k=41)+TEMP+ RH + as.factor(dow), family=quasipoisson, data=hou_pm, method="REML")
```

::::: side-by-side
::: text
When $\text{PM}_{2.5}$ was fit with a penalized spline for the all-cause and respiratory mortality models, the estimated degrees of freedom was 1 for each (Supplemental). However, the model for CVD mortality showed an estimated df of 2.18, signifying a non-linear relationship (Figure 2). The dose-response curves for the all-cause mortality and respiratory outcomes show a linear, positive relationship $\text{PM}_{2.5}$ concentrations, with respiratory disease having larger relative risk estimates as $\text{PM}_{2.5}$ increases. For the $\text{PM}_{2.5}$-CVD mortality plot, the log rate ratio peaks around 17 $\mu g/m^3$, followed by a gradual decrease at higher concentrations. The confidence intervals for all-cause and CVD mortality are wide, with the former's lower boundary crossing 1. Due to the non-linear relationship between CVD and $\text{PM}_{2.5}$, it would not be appropriate to calculate relative risk. A 10 $\mu g/m^3$ increase in $\text{PM}_{2.5}$ was associated with a 1% increase in all-cause mortality (RR=1.01, 95% CI: 0.999, 1.021) 5% increase in respiratory mortality (RR=1.05, 95% CI: 1.012, 1.092) [Table 2].
:::

::: table
```{r pm_10inc, include=FALSE}
b_tot<-mod_tot$coef["pm25"]
se_tot<-sqrt(vcov(mod_tot)["pm25","pm25"])

exp(b_tot*10)
exp((b_tot -1.96*se_tot)*10)
exp((b_tot +1.96*se_tot)*10)

b_resp<-mod_resp$coef["pm25"]
se_resp<-sqrt(vcov(mod_resp)["pm25","pm25"])

exp(b_resp)
exp((b_resp -1.96*se_resp)*10)
exp((b_resp +1.96*se_resp)*10)
```

```{r table2, echo=FALSE}
table2_vars <- data.frame(var=c("Coefficient (exp)","p-value", 'RR per 10 µg/m³ (95% CI)'))
table2_clean <-  data.frame(Measure = c("Exp. Estimate", "p-value","RR per 10 µg/m³ (95% CI)"),
  tot_mort = c("0.001","0.0005","1.01 (0.999, 1.021)"),
  resp_mort = c("0.005","0.0019","1.05 (1.012, 1.092)")
)
table2_clean %>% gt() %>%
  tab_header(title = md("**Table 2. Coefficient p-value for PM2.5 Models Where Exposure-Response Relationship was Linear**"),
    subtitle = "Relative risk per 10 µg/m³ increase"
  ) %>%

  cols_label( Measure="Measure",
    tot_mort="PM2.5 (all-cause mortality)",
    resp_mort="PM2.5 (respiratory mortality)" 
    ) %>%
  cols_align("left", columns = Measure) %>%
  cols_align("center", columns = -Measure)
```
:::
:::::

Testing the significance of temperature as a covariate, the ANOVA result for the all-cause mortality model showed a p-value below 0.05 (p-value=0.0005; residual deviance of 2583.8 and 2571.5). However, the results for CVD and respiratory mortality showed large p-values (p-values of 0.38 and 0.48, respectively). Fitting a penalized spline to temperature resulted in estimated degrees of freedom of 1 for all three of our outcomes (Supplemental). The refitted dose-response curves show a positive, linear relationship between temperature and daily mortality (Figure 3). However, the confidence intervals of the relative risk for each model is wide, crossing 1 for both the CVD and respiratory mortality models where temperatures are probable (e.g, mortality temperature below -20 might not provide relevant public health information to the citizens of Houston). All-cause mortality maintained an estimated relative risk above 1 for each observed temperature.

```{r doseresponse_pm25, echo=FALSE}
par(mfrow = c(1, 3), oma = c(4, 5, 4, 1), mar = c(3, 4, 2, 1))
#all-cause
predpm25<-predict(mod_tot,type='terms',se.fit=TRUE)
predpm25.se<-predpm25$se.fit

rr_tot<-(exp(predpm25$fit[,"pm25"]))
cihi_tot <- (exp((predpm25$fit[,"pm25"] +1.96*predpm25.se[,"pm25"])))
cilo_tot<- (exp((predpm25$fit[,"pm25"] -1.96* predpm25.se[,"pm25"])))

plot(hou_pm$pm25,rr_tot,pch=17,ylim=c(min(cilo_tot),max(cihi_tot)),
         ylab="Relative Risk ", main="All-Cause Mortality", cex=.8)
points(hou_pm$pm25,cihi_tot,pch=18,cex=.5)
points(hou_pm$pm25,cilo_tot,pch=18,cex=.5)

#resp
predpm25<-predict(mod_resp,type='terms',se.fit=TRUE)
predpm25.se<-predpm25$se.fit
rr_resp<-(exp(predpm25$fit[,"pm25"]))
cihi_resp <- (exp((predpm25$fit[,"pm25"] +1.96*predpm25.se[,"pm25"])))
cilo_resp<- (exp((predpm25$fit[,"pm25"] -1.96* predpm25.se[,"pm25"])))

plot(hou_pm$pm25,rr_resp,pch=17,ylim=c(min(cilo_resp),max(cihi_resp)),
         ylab="Relative Risk ", main="Respiratory Mortality", cex=.8)
points(hou_pm$pm25,cihi_resp,pch=18,cex=.5)
points(hou_pm$pm25,cilo_resp,pch=18,cex=.5)


#CVD
plot(mod_cvd,  select=1, scale=0, main="Cardiovascular Mortality", ylab = "Log Rate Ratio (edf = 2.18")

figure2title<-'Figure 2. Dose-Response Curves for Same-Day PM2.5 and Daily Mortality, Houston 1999-2006'
title_f2 <- paste(strwrap(figure2title, width = 70), collapse = "\n")

mtext(title_f2, side = 3, line = 1, outer = TRUE, cex = 1, font = 2)
mtext("PM2.5 ", side = 1, line = 1, outer = TRUE, cex = 1.0, font = 2)
```

```{r temperature_inc, echo=FALSE}
mod_tot_notemp<-gam(tot ~ s(pm25)+s(time,bs = "cr",fx=TRUE,k=41)+as.factor(dow)+ RH , family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML") 
mod_cvd_notemp<-gam(cvd ~ s(pm25)+s(time,bs = "cr",fx=TRUE,k=41)+as.factor(dow)+ RH, family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML") 
mod_resp_notemp<-gam(resp ~ s(pm25)+s(time,bs = "cr",fx=TRUE,k=41)+as.factor(dow)+ RH, family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML")

#anova(mod_tot_notemp, mod_tot, test = "F")
#anova(mod_cvd_notemp, mod_cvd, test = "F")
#anova(mod_resp_notemp, mod_resp, test = "F")
```

```{r doseresponse_temp, echo=FALSE}
par(mfrow = c(1, 3), oma = c(4, 5, 4, 1), mar = c(3, 4, 2, 1))
#all-cause
predTEMP<-predict(mod_tot,type='terms',se.fit=TRUE)
predTEMP.se<-predTEMP$se.fit

rr_tot<-(exp(predTEMP$fit[,"TEMP"]))
cihi_tot <- (exp((predTEMP$fit[,"TEMP"] +1.96*predTEMP.se[,"TEMP"])))
cilo_tot<- (exp((predTEMP$fit[,"TEMP"] -1.96* predTEMP.se[,"TEMP"])))

plot(hou_pm$TEMP,rr_tot,pch=17,ylim=c(min(cilo_tot),max(cihi_tot)),
         ylab="Relative Risk ", main="All-Cause Mortality", cex=.8)
points(hou_pm$TEMP,cihi_tot,pch=18,cex=.5)
points(hou_pm$TEMP,cilo_tot,pch=18,cex=.5)

#resp
predTEMP<-predict(mod_resp,type='terms',se.fit=TRUE)
predTEMP.se<-predTEMP$se.fit

rr_resp<-(exp(predTEMP$fit[,"TEMP"]))
cihi_resp <- (exp((predTEMP$fit[,"TEMP"] +1.96*predTEMP.se[,"TEMP"])))
cilo_resp<- (exp((predTEMP$fit[,"TEMP"] -1.96* predTEMP.se[,"TEMP"])))

plot(hou_pm$TEMP,rr_resp,pch=17,ylim=c(min(cilo_resp),max(cihi_resp)),
         ylab="Relative Risk ", main="Respiratory Mortality", cex=.8)
points(hou_pm$TEMP,cihi_resp,pch=18,cex=.5)
points(hou_pm$TEMP,cilo_resp,pch=18,cex=.5)


#CVD
predTEMP<-predict(mod_cvd,type='terms',se.fit=TRUE)
predTEMP.se<-predTEMP$se.fit
rr_cvd<-(exp(predTEMP$fit[,"TEMP"]))
cihi_cvd <- (exp((predTEMP$fit[,"TEMP"] +1.96*predTEMP.se[,"TEMP"])))
cilo_cvd<- (exp((predTEMP$fit[,"TEMP"] -1.96* predTEMP.se[,"TEMP"])))

plot(hou_pm$TEMP,rr_cvd,pch=17,ylim=c(min(cilo_cvd),max(cihi_cvd)),
         ylab="Relative Risk ", main="CVD Mortality", cex=.8)
points(hou_pm$TEMP,cihi_cvd,pch=18,cex=.5)
points(hou_pm$TEMP,cilo_cvd,pch=18,cex=.5)

figure3title<-'Figure 3. Dose-Response Curves for Same-Day Temperature and Daily Mortality, Houston 1999-2006'
title_f3 <- paste(strwrap(figure3title, width = 70), collapse = "\n")
mtext(title_f3, side = 3, line = 1, outer = TRUE, cex = 1, font = 2)
mtext("Temperature (in Farenheit)", side = 1, line = 1, outer = TRUE, cex = 1.0, font = 2)
```

### Discussion

These results on the association between same-day $\text{PM}_{2.5}$ and daily mortality provide some evidence that there is a significant relationship between the two. However, significance seems only to apply to daily mortality of a certain kind. The dose-response curve for $\text{PM}_{2.5}$ and respiratory mortality had only positive relative risk estimates and the confidence bands did not cross 1. Unlike the other two outcomes, this provides evidence that there is a significant, positive, and linear relationship between $\text{PM}_{2.5}$ and respiratory mortality. Additional evidence of this can be seen in the confidence interval for a 10-unit increase of $\text{PM}_{2.5}$, as the interval doesn't contain 1.

This study was subject to many limitations. Although the ANOVA results for two of the outcomes indicated temperature should not be included in the model, the choice was made to include it. Previous studies have noted temperature as a confounder between $\text{PM}_{2.5}$ and mortality. Even though adjusting for time and RH might have accounted for most of the possible confounding, adjusting for temperature might reduce residual confounding. There are many ways to fit splines to capture trends and seasonality. One option is to fut a penalized spline using restricted maximum likelihood, though through this method there may be potential to over-smooth the data. In our analysis, we used a natural spline with 4 df per year (40 total degrees of freedom) as this prevents any over smoothing, and is more interpretable. However, one disadvantage to our method is that the choice of df is arbitrary and it prevents the model to be flexible and adapt to the structure of the data. Finally, this study did not account for the lagged effect of $\text{PM}_{2.5}$ concentrations. The lagged effect of $\text{PM}_{2.5}$ might be significant driver in mortality, but that was not explored in this analysis.

Overall, respiratory and same-day $\text{PM}_{2.5}$ in Houston might have a moderate linear relationship. Reduction in emissions might prove to lower overall respiratory mortality. Further studies are needed to assess the lagged effect of $\text{PM}_{2.5}$ and different forms of mortality. Additionally, studies regarding $\text{PM}_{2.5}$ and it's effect on other outcomes (e.g., neuro-cognitive disorders) will create a better public health understanding of the impacts of this pollutant.

### References

::: hanging-indent
<p>

1.  EPA. Particulate Matter (PM) Basics. <https://www.epa.gov/pm-pollution/particulate-matter-pm-basics>. Last updated May 2025.

    </p>

<p>

2.  Chao, C. Y., Li, W., Hopke, P. K., Guo, F., Wang, Y., & Griffin, R. J. (2025). Increases in PM2.5 levels in Houston are associated with a highly recirculating sea breeze. Environmental pollution (Barking, Essex : 1987), 366, 125381. <https://doi.org/10.1016/j.envpol.2024.125381>

    </p>

    <a > </a>

### Supplemental

```{r modeltests, echo=FALSE}
model_fit<-function(outcome) {
###TEMP Spline
  form_temp_sp<-as.formula(paste(outcome, "~ s(time,bs =   'cr',fx=TRUE,k=41)+s(TEMP)+as.factor(dow)+ RH +  s(pm25)"))
    mod_temp_sp_g<- gam(form_temp_sp, family=quasipoisson, data=hou_pm, na.action=na.omit)
   mod_temp_sp_r<- gam(form_temp_sp, family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML")
   
  temp_sum<-summary(mod_temp_sp_g)$s.table
  temp_edf_g<-temp_sum["s(TEMP)","edf"]
  temp_p_g<-temp_sum["s(TEMP)","p-value"]
  
  temp_sum<-summary(mod_temp_sp_r)$s.table
  temp_edf_r<-temp_sum["s(TEMP)","edf"]
  temp_p_r<-temp_sum["s(TEMP)","p-value"]
  
###PM2.5 Spline 
   form_pm25_sp<-as.formula(paste(outcome,"~ s(time,bs =   'cr',fx=TRUE,k=41)+TEMP+as.factor(dow)+ RH +  s(pm25)"))
   mod_pm25_sp_g<- gam(form_pm25_sp, family=quasipoisson, data=hou_pm, na.action=na.omit)
   mod_pm25_sp_r<- gam(form_pm25_sp, family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML")
   
  pm25_sum<-summary(mod_pm25_sp_g)$s.table
  pm25_edf_g<-pm25_sum["s(pm25)","edf"]
  pm25_p_g<-pm25_sum["s(pm25)","p-value"]
  
    pm25_sum<-summary(mod_pm25_sp_r)$s.table
  pm25_edf_r<-pm25_sum["s(pm25)","edf"]
  pm25_p_r<-pm25_sum["s(pm25)","p-value"]
  
###RH Spline 
   form_RH_sp<-as.formula(paste(outcome,"~ s(time,bs =   'cr',fx=TRUE,k=41)+TEMP+as.factor(dow)+ s(RH) +  s(pm25)"))
    mod_RH_sp_g<- gam(form_RH_sp, family=quasipoisson, data=hou_pm, na.action=na.omit)
   mod_RH_sp_r<- gam(form_RH_sp, family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML")
  RH_sum<-summary(mod_RH_sp_g)$s.table
  RH_edf_g<-RH_sum["s(RH)","edf"]
  RH_p_g<-RH_sum["s(RH)","p-value"] 
  
  RH_sum<-summary(mod_RH_sp_r)$s.table
  RH_edf_r<-RH_sum["s(RH)","edf"]
  RH_p_r<-RH_sum["s(RH)","p-value"]
  
 results<-data.frame(
    edf_=c(temp_edf_g, pm25_edf_g, RH_edf_g, temp_edf_r, pm25_edf_r, RH_edf_r),
    p_val_=c(temp_p_g, pm25_p_g, RH_p_g,temp_p_r, pm25_p_r, RH_p_r)
  )
colnames(results) <- paste0(colnames(results), outcome)
assign(paste0("edf_", outcome), results, envir =.GlobalEnv)
}
model_fit("tot")
model_fit("cvd")
model_fit("resp")
edf_vars<-data.frame(
var=c("TEMP_GCV", "PM2.5_GCV", "RH_GCV", "TEMP_REML", "PM2.5_REML", "RH_REML"))
edf_all<-cbind(edf_vars,edf_tot,edf_cvd,edf_resp)
edf_all
```

```{r models_allspline, include=TRUE}
mod_tot<-gam(tot ~  s(pm25) + s(time,bs = "cr",fx=TRUE,k=41)+TEMP+ RH + as.factor(dow), family=quasipoisson, data=hou_pm, method="REML")


mod_cvd<-gam(cvd ~ s(pm25) + s(time,bs = "cr",fx=TRUE,k=41)+TEMP+ RH + as.factor(dow), family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML")

mod_resp<-gam(resp ~ pm25 + s(time,bs = "cr",fx=TRUE,k=41)+TEMP+ RH + as.factor(dow), family=quasipoisson, data=hou_pm, method="REML")
```

```{r doseresponse_allsplines}
par(mfrow=c(1,3))
plot(mod_tot,  select=1, scale=0, main="Total Mortality")
plot(mod_cvd,  select=1, scale=0, main="Cardiovascular Mortality")
plot(mod_resp, select=1, scale=0, main="Respiratory Mortality",)
```
:::
