---
title: "Association of Same-Day Ambient Fine Particulate Matter with Mortality"
author: "Anton Roche, Grace Gaultier, and Joel Chavez Ortiz"
date: "ID 271 - Midterm"
output: 
  html_document:
    toc_float: true
    toc: true
---

```{css, echo=FALSE}
.hanging-indent p {
  padding-left: 2em;
  text-indent: -2em;}
.side-by-side {
  display: flex;
  gap: 40px;
  align-items: flex-start;
}
.side-by-side .text {
  flex: 1;
}
.side-by-side .table {
  flex: 1.2;
}
```

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "~/Desktop/Harvard/Courses/Spring 2026/ID 271")
```

```{r packages, message=F, include=FALSE}
library(foreign)
library(stats)
library(mgcv)
library(splines)
library(nlme)
library(table1)
library(lubridate)
library(gplots)
library(MASS)
```

```{r data, include=FALSE}
hou_pm <- read.csv("hou_pm.csv")
```

### Background

Emissions from both natural and anthropogenic sources can create particulate matter (PM) small enough to contribute to adverse health effects in humans (e.g., damage to lung function, asthma) and contribute to environmental damage (e.g., acid rain effects, disruption of nutrient balance in soil and rivers)$\text{.}^{1}$) PM with a diameter of 2.5 micrometers or smaller $\text{PM}_{2.5}$ are considered fine inhalable particles that can penetrate deeper into the lungs than larger PM. Houston is a large U.S. city with a large urban population. High PM emissions from vehicles, industry, and other sources may pose a risk to population health. An observational time-series analysis was conducted to further understand the association between same-day ambient concentrations of $\text{PM}_{2.5}$ and daily mortality.

### Methods

Using historical records data from Houston between January 1, 1999 and December 31, 2006, daily deaths, $\text{PM}_{2.5}$ concentration, and meteorlogic were extracted to analyze the association between $\text{PM}_{2.5}$ and three outcomes of interest: all-cause mortality, cardiovascular (CVD) disease-related mortality, and respiratory disease-related mortality.

#### Study Analysis

The association between same-day concentrations of $\text{PM}_{2.5}$ and the three outcomes of interest was estimated fitting generalized additive models (GAM) with a quasi-Poisson regression. When possible, relative risk (RR) was estimated between $\text{PM}_{2.5}$ and our outcomes if they seemed to have a linear relationship. A quasi-Poisson model provides an advantage over a standard Poisson regression in that it allows for data to be overdispersed. The adjusted GAMs included relative humidity (RH), day of the week (dow), average daily temperature (TEMP), and days since the start of the study (time) as relevant covariates. Time was modeled with a natural spline using 5 degrees of freedom for each of the 8 years of the study (40 degrees of freedom) to account for trends in seasonality and variation between years. Penalized splines were initially fit for relative humidity (RH) and temperature. RH and temperature were found to have linear in the penalized regression, so the model was refit with RH and temperature as linear predictors (dose response figures). Splines were fitted with cubic regression smoothing functions generated from restricted maximum likelihood (REML) for more stable parameter estimation and less overfitting.

$$\text{log}(\mathbb{E}[Y_{t}]) = \beta_{0} +f_{1}(PM_{t})+f_{2}(Time_{t})+\beta_{1}Temp_{t}+\beta_{2}RH_{t}+\beta_{3}Dow_{t}$$There $Y_t$ is the count of mortality (total, CVD, or respiratory) on day t since the beginning of the study, RH is the same day average relative humidity (%), Temp is the same day average temperature (Fahrenheit), $\text{PM}_{2.5}$ is the same day average fine particulate matter concentration (in $\mu g/m^3$), and dow is the day of the week as a categorical variable defined from 1 (Monday) through 7 (Sunday).

<div class="side-by-side">

<div class="text">

### Results

The mean for daily all-cause, CVD, and respiratory mortality were 50.08 (variance: 62.28, SD: 7.89), 14.89 (variance: 16.56, SD: 4.07), and 3.99 (variance: 4.67, SD: 2.16), respectively (Table 1). Other the study period, there was a total of 146,344 recorded deaths from any cause. Of these deaths, 29.7% (43,523/145,344) were from CVD and 8% (11,657/145,344) were from respiratory disease. The median concentration of $\text{PM}_{2.5}$ was 12 $\mu g/m^3$, with the highest single day concentration being 53 $\mu g/m^3$. Average temperature over the 8 year period was 70.2°  with average relative humidity being 73.17%.

</div>

<div class="table">
```{r table1, echo=FALSE}
label(hou_pm$tot)  <- "All-cause mortality, daily"
label(hou_pm$cvd)  <- "Cardiovascular mortality, daily"
label(hou_pm$resp) <- "Respiratory mortality, daily"
label(hou_pm$pm25) <- "PM2.5, µg/m³"
label(hou_pm$TEMP) <- "Temperature, °F"
label(hou_pm$RH)   <- "Relative Humidity, %"

table1(~ tot + cvd + resp + pm25 + TEMP + RH,  data = hou_pm, caption = "<span style='color: black; font-weight: bold;'>Table 1. Descriptive Statistics of Daily Mortality, Daily Particulate Matter, and Weather, Houston 1999-2006", overall = "Descriptive Statistic Values</span>",
       render.continuous = function(x) {
         outc <- attr(x, "label")
         if (outc %in% c("All-cause mortality, daily", "Cardiovascular mortality, daily","Respiratory mortality, daily")) {
        c(" " = "", "Mean (SD)" = sprintf("%.2f (%.2f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE)),
           "Variance" = sprintf("%.2f", var(x, na.rm = TRUE)),
           
           "Min, Median, Max" = sprintf("%.0f, %.0f, %.0f", min(x, na.rm = TRUE),  median(x, na.rm = TRUE),  max(x, na.rm = TRUE)), "~Total Deaths, 1999-2006" = format(sum(x, na.rm = TRUE), big.mark=","))                            
         }
        else {
            c(" " = "", "Mean (SD)" = sprintf("%.2f (%.2f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE)),
            "Min, Median, Max" = sprintf("%.0f, %.0f, %.0f", min(x, na.rm = TRUE),  median(x, na.rm = TRUE),  max(x, na.rm = TRUE))) 
        }
       })

```

</div>
</div>
```{r, echo=FALSE}
par(mfrow = c(1, 3), oma = c(4, 4, 3, 0), mar = c(2, 2, 2, 1)) 
hist(hou_pm$tot, main ="All-Cause Mortality",)
hist(hou_pm$cvd,main ="CVD Mortality")
hist(hou_pm$resp, main ="Respiratory Mortality")
mtext("Figure 1. Histograms of Daily Mortality, Houston 1999-2006", side = 3, line = 1, outer = TRUE, cex = 1.5, font = 2)
mtext("Number of Deaths ", side = 2, line = 1, outer = TRUE, cex = 1.0, font = 2)
mtext("Frequency", side = 1, line = 1, outer = TRUE, cex = 1.0, font = 2)
```



### References

::: hanging-indent
<p>

1.  EPA. Particulate Matter (PM) Basics. <https://www.epa.gov/pm-pollution/particulate-matter-pm-basics>. Last updated May 2025.

    </p>

<p>

2.  Chao, C. Y., Li, W., Hopke, P. K., Guo, F., Wang, Y., & Griffin, R. J. (2025). Increases in PM2.5 levels in Houston are associated with a highly recirculating sea breeze. Environmental pollution (Barking, Essex : 1987), 366, 125381. <https://doi.org/10.1016/j.envpol.2024.125381>

    </p>


