---
title: "ID271 Midterm"
author: "JCO"
date: "Due Feb 22, 2026 at 11:59 PM (EST)"
output: 
  html_document:
    toc_float: true
    toc: true
---
# Directory
``` {r setup}
knitr::opts_knit$set(root.dir = "~/Desktop/Harvard/Courses/Spring 2026/ID 271")
```

# Packages
```{r, message=F}
library(foreign)
library(stats)
library(mgcv)
library(splines)
library(nlme)
library(lubridate)
library(gplots)
library(MASS)
```

```{r}
hou_pm <- read.csv("hou_pm.csv")
```

```{r}
names(hou_pm)
summary(hou_pm)
```

# Check to see if mean=variance
```{r}
hist(hou_pm$tot)
hist(hou_pm$cvd)
hist(hou_pm$cvd)
summary(hou_pm$tot)
mean(hou_pm$tot)
var(hou_pm$tot)
```
```{r}
ll <- matrix(nrow=4,ncol=6)
ll[1,] <- rep(1,6) #intercept
ll[2,] <- 0:5 #lag0 to lag5, i term
ll[3,] <- ll[2,]^2 #i^2 term
ll[4,] <- ll[2,]^3 #i^3 term

ll <- t(ll) #transpose
```

# Construct Z matrix; create PM2.5 and temp lags
```{r}
lagm<-function(x,n) {
  nn<-length(x)
  xn<-c(rep(NA,n),x[1:(nn-n)])
  return(xn)
}

lenpol <- length(hou_pm[,1])
xx<-matrix(nrow=lenpol,ncol=6)
for (i in 1:6) {
	xx[,i]<-lagm(hou_pm$pm25,(i-1))
}

zz<-xx %*% ll

colnames(xx) <- paste0("pm25", 1:6)
hou_pm <- cbind(hou_pm, xx)

#make month variable
hou_pm$date <-as.Date(hou_pm$date, format="%m/%d/%Y")
hou_pm$mon<-month(hou_pm$date)

#make time variable
hou_pm$time<-1:dim(hou_pm)[1]
hou_pm$temp1=lagm(hou_pm$TEMP,1)

cor(hou_pm$TEMP[!is.na(hou_pm$temp1)],hou_pm$temp1[!is.na(hou_pm$temp1)])
```

```{r}
try<-gam(tot~s(TEMP)+ s(temp1)+s(time)+s(RH) +as.factor(mon),data=hou_pm,method="REML")
plot(try)
AIC(try)

try2<-gam(tot~s(TEMP,temp1)+s(time)+s(RH) +as.factor(mon),data=hou_pm,method="REML")
plot(try2)
AIC(try2)
```

```{r}
mod<-gam(tot ~ s(time,bs = "cr",fx=FALSE,k=41)+s(TEMP,bs = "cr",fx=TRUE,k=5)+s(temp1)+as.factor(dow) + RH + zz[,1] + zz[,2] + zz[,3] +zz[,4], data=hou_pm,method="REML", na.action=na.omit)
summary(mod)

```

How do we get effect estimates at each lag? Remember the equation bi=c0 +c1i +c2i^2 +c3i^3. The coefficients of the Z's are the c's. So we will need to grab them and multiply them by the appropriate weights for each i (that is, by 1 for i=1, etc)

```{r}
#obtain the coef. There are 8 variables that precede 
# the Z's--intercept, 6 dow dummies, and RH
#So the Z's are the 9-12th coefficients

pm25_est_dl <- as.matrix(mod$coef[21:24]) #c matrix
#now multiple them by the weights to become b's
pm25_est <- ll%*%as.matrix(pm25_est_dl) #beta matrix

#obtain the se of the b's

#the b's are sums of the c's, so we need the covariance of the c's to compute the se of the b's. Vcov will give us the covariance matrix of all coefficents. We subset it the the 9:12 by 9:12 part for just the covariance of the b's

pm25_cov <- ll%*%as.matrix(vcov(mod)[21:24,21:24])%*%t(ll)

pm25_se <- sqrt(diag(pm25_cov))

#build the effect estimates matrix for beta
pm25_est_se <- data.frame(var=c('pm25_lag5','pm25_lag4','pm25_lag3','pm25_lag2','pm25_lag1','pm25_lag0'),
                          coef=pm25_est,
                          se=pm25_se)
```

####plot
```{r}
lags<-c(5,4,3,2,1,0)
beta=pm25_est_se[,2]
se=pm25_est_se[,3]
plotCI(y=beta,x=lags,uiw=1.96*se,err="y")
```

We can compare this to the unconstrained effects.  

```{r}
#unconstrained fitting
mod2<-gam(tot ~ s(time,bs = "cr",fx=TRUE,k=40)+s(TEMP)+s(temp1)
          +as.factor(dow)+RH+pm25+xx[,2]+xx[,3]+xx[,4]+xx[,5]+xx[,6], 
          family=quasipoisson, method="REML",data=hou_pm, na.action=na.omit)

#plot
lags<-c(0,1,2,3,4,5)
beta<-c(mod2$coef[9],mod2$coef[10],mod2$coef[11],mod2$coef[12],mod2$coef[13],mod2$coef[14])
plotCI(y=beta,x=lags,uiw=1.96*sqrt(diag(mod2$Vp[9:14,9:14])),err="y")
```
# Dose-response curves
```{r}
mod_tot<-gam(tot ~ s(time,bs = "cr",fx=TRUE,k=41)+s(TEMP,bs = "cr",fx=TRUE,k=4)+as.factor(dow)+ RH +  s(pm25), family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML")

mod_cvd<-gam(cvd ~ s(time,bs = "cr",fx=TRUE,k=41)+TEMP+as.factor(dow)+ RH +  s(pm25), family=quasipoisson, data=hou_pm, na.action=na.omit)

mod_resp<-gam(resp ~ s(time,bs = "cr",fx=TRUE,k=41)+s(TEMP,bs = "cr",fx=TRUE,k=5)+as.factor(dow)+ RH +  s(pm25), family=quasipoisson, data=hou_pm, na.action=na.omit ,method="REML")

summary(mod1)
model<-gam(tot ~ s(time,bs = "cr",k=41)+s(TEMP,bs = "cr",k=5)+as.factor(dow)+ RH +  s(pm25), family=quasipoisson, data=hou_pm,method="REML", na.action=na.omit)

plot(mod_tot,scale=0,pages=1, main="total")
plot(mod_cvd,scale=0,pages=1, main="cvd")
plot(mod_resp,scale=0,pages=1, main="resp")
plot(model,scale=0,pages=1)
```

