---
title: "ID271 Midterm"
author: "JCO"
date: "Due Feb 22, 2026 at 11:59 PM (EST)"
output: 
  html_document:
    toc_float: true
    toc: true
---

### Methods

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "~/Desktop/Harvard/Courses/Spring 2026/ID 271")
```

# Packages

```{r, message=F, include=FALSE}
library(foreign)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(stats)
library(mgcv)
library(table1)
library(splines)
library(nlme)
library(gt)
library(lubridate)
library(gplots)
library(MASS)
```

```{r, include=FALSE}
hou_pm <- read.csv("hou_pm.csv")
```

```{r, include=FALSE}
names(hou_pm)
sum(hou_pm$tot)
sum(hou_pm$cvd)
sum(hou_pm$resp)
summary(hou_pm$pm25)
```

# Check to see if mean=variance



summary(hou_pm$tot)
mean(hou_pm$tot)
var(hou_pm$tot)


summary(hou_pm$cvd)
mean(hou_pm$cvd)
var(hou_pm$cvd)

summary(hou_pm$resp)
mean(hou_pm$resp)
var(hou_pm$resp)
```

```{r table1}
label(hou_pm$tot)  <- "All-cause mortality (mean [SD])"
label(hou_pm$cvd)  <- "Cardiovascular mortality (mean [SD])"
label(hou_pm$resp) <- "Respiratory mortality (mean [SD])"
label(hou_pm$pm25) <- "PM2.5 (µg/m³)"
label(hou_pm$TEMP) <- "Temperature (°F)"
label(hou_pm$RH)   <- "Relative Humidity (%)"

table1(~ tot + cvd + resp + pm25 + TEMP + RH,  data = hou_pm, overall = "Descriptive Statistics",
       render.continuous = function(x) {
         c("Mean [SD]" = sprintf("%.2f (%.2f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE)),
           "Min, 25th, Median, 75th, Max" = sprintf("%.2f, %.2f, %.2f, %.2f, %.2f",
                                                      min(x, na.rm = TRUE),
                                                      quantile(x, 0.25, na.rm = TRUE),
                                                      median(x, na.rm = TRUE),
                                                      quantile(x, 0.75, na.rm = TRUE),
                                                      max(x, na.rm = TRUE)))
       })
```

```{r}
colnames(xx) <- paste0("pm25", 1:6)
hou_pm <- cbind(hou_pm, xx)

#make month variable
hou_pm$date <-as.Date(hou_pm$date, format="%m/%d/%Y")
hou_pm$mon<-month(hou_pm$date)

#make time variable
hou_pm$time<-1:dim(hou_pm)[1]
hou_pm$temp1=lagm(hou_pm$TEMP,1)

cor(hou_pm$TEMP[!is.na(hou_pm$temp1)],hou_pm$temp1[!is.na(hou_pm$temp1)])
```

```{r modeltests}
model_fit<-function(outcome) {
###TEMP Spline
  form_temp_sp<-as.formula(paste(outcome, "~ s(time,bs =   'cr',fx=TRUE,k=41)+s(TEMP)+as.factor(dow)+ RH +  s(pm25)"))
    mod_temp_sp_g<- gam(form_temp_sp, family=quasipoisson, data=hou_pm, na.action=na.omit)
   mod_temp_sp_r<- gam(form_temp_sp, family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML")
   
  temp_sum<-summary(mod_temp_sp_g)$s.table
  temp_edf_g<-temp_sum["s(TEMP)","edf"]
  temp_p_g<-temp_sum["s(TEMP)","p-value"]
  
  temp_sum<-summary(mod_temp_sp_r)$s.table
  temp_edf_r<-temp_sum["s(TEMP)","edf"]
  temp_p_r<-temp_sum["s(TEMP)","p-value"]
  
###PM2.5 Spline 
   form_pm25_sp<-as.formula(paste(outcome,"~ s(time,bs =   'cr',fx=TRUE,k=41)+TEMP+as.factor(dow)+ RH +  s(pm25)"))
   mod_pm25_sp_g<- gam(form_pm25_sp, family=quasipoisson, data=hou_pm, na.action=na.omit)
   mod_pm25_sp_r<- gam(form_pm25_sp, family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML")
   
  pm25_sum<-summary(mod_pm25_sp_g)$s.table
  pm25_edf_g<-pm25_sum["s(pm25)","edf"]
  pm25_p_g<-pm25_sum["s(pm25)","p-value"]
  
    pm25_sum<-summary(mod_pm25_sp_r)$s.table
  pm25_edf_r<-pm25_sum["s(pm25)","edf"]
  pm25_p_r<-pm25_sum["s(pm25)","p-value"]
  
###RH Spline 
   form_RH_sp<-as.formula(paste(outcome,"~ s(time,bs =   'cr',fx=TRUE,k=41)+TEMP+as.factor(dow)+ s(RH) +  s(pm25)"))
    mod_RH_sp_g<- gam(form_RH_sp, family=quasipoisson, data=hou_pm, na.action=na.omit)
   mod_RH_sp_r<- gam(form_RH_sp, family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML")
  RH_sum<-summary(mod_RH_sp_g)$s.table
  RH_edf_g<-RH_sum["s(RH)","edf"]
  RH_p_g<-RH_sum["s(RH)","p-value"] 
  
  RH_sum<-summary(mod_RH_sp_r)$s.table
  RH_edf_r<-RH_sum["s(RH)","edf"]
  RH_p_r<-RH_sum["s(RH)","p-value"]
  
 results<-data.frame(
    edf_=c(temp_edf_g, pm25_edf_g, RH_edf_g, temp_edf_r, pm25_edf_r, RH_edf_r),
    p_val_=c(temp_p_g, pm25_p_g, RH_p_g,temp_p_r, pm25_p_r, RH_p_r)
  )
colnames(results) <- paste0(colnames(results), outcome)
assign(paste0("edf_", outcome), results, envir =.GlobalEnv)
}
model_fit("tot")
model_fit("cvd")
model_fit("resp")
edf_vars<-data.frame(
var=c("TEMP_GCV", "PM2.5_GCV", "RH_GCV", "TEMP_REML", "PM2.5_REML", "RH_REML"))
edf_all<-cbind(edf_vars,edf_tot,edf_cvd,edf_resp)

mod_tot<-gam(tot ~ s(pm25)+s(time,bs = "cr",fx=TRUE,k=41)+s(TEMP,bs = "cr",fx=TRUE,k=4)+as.factor(dow)+ RH , family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML")
plot(try)


try2<-gam(tot~s(TEMP,temp1)+s(time)+s(RH) +as.factor(mon),data=hou_pm,method="REML")
plot(try2)

```

```{r}
mod_tot<-gam(tot ~ s(pm25)+s(time,bs = "cr",fx=TRUE,k=41)+TEMP+as.factor(dow)+ RH , family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML") 
summary(mod_tot)$dispersion
mod_cvd<-gam(cvd ~ s(pm25)+s(time,bs = "cr",fx=TRUE,k=41)+TEMP+as.factor(dow)+ RH, family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML") 
summary(mod_cvd)
mod_resp<-gam(resp ~ s(pm25)+s(time,bs = "cr",fx=TRUE,k=41)+TEMP+as.factor(dow)+ RH, family=quasipoisson, data=hou_pm, na.action=na.omit, method="REML")
summary(mod_resp)$dispersion
beta_tot <- coef(mod_tot)["pm25"]
beta_resp <- coef(mod_resp)["pm25"]


#se<-sqrt(mod1$Vp[9,9])

#exp(beta*10)
#exp((beta -1.96*se)*10)
#exp((beta +1.96*se)*10)
```

# Dose-response curves

```{r}
mod_tot<-gam(tot ~ s(time,bs = "cr",fx=TRUE,k=41)+s(TEMP,bs = "cr",fx=TRUE,k=4)+as.factor(dow)+ RH +  s(pm25), family=quasipoisson, data=hou_pm, method="REML")

mod_cvd<-gam(cvd ~ s(time,bs = "cr",fx=TRUE,k=41)+TEMP+as.factor(dow)+ RH +  s(pm25), family=quasipoisson, data=hou_pm, na.action=na.omit)

mod_resp<-gam(resp ~ s(time,bs = "cr",fx=TRUE,k=41)+TEMP+as.factor(dow)+ RH +  s(pm25), family=quasipoisson, data=hou_pm ,method="REML")


model<-gam(tot ~ s(time,bs = "cr",k=41)+s(TEMP,bs = "cr",k=5)+as.factor(dow)+ RH +  s(pm25), family=quasipoisson, data=hou_pm,method="REML")

plot(mod_tot,scale=0,pages=1, main="total")
plot(mod_cvd,scale=0,pages=1, main="cvd")
plot(mod_resp,scale=0,pages=1, main="resp")
plot(model,scale=0,pages=1)
```
